---
title: "September 11th Terrorist Network Analysis"
author: "Marc Petta"
date:
output:
  html_document: default
  word_document: default
---

## Introduction 
The analysis to follow seeks to determine characteristics of the terrorist network that was responsible for the attacks on the United States on September 11th, 2001. Additionally, a section is included which attempts to model the network to determine if predictions can be made based on the modeled data.

### Load data and output csv of vertices and edges
```{r, message=FALSE}

# repeated calls to libraries occur throughout to avoid dependency conflicts
library(statnet) 

# load the rda file
load(file = "Krebs.rda")
# supplement existing dataset
Krebs %v% "Age" = c(24,29,26,20,25,23,21,20,26,22,20,22,24,23,33,22,22,28,25)
Krebs %v% "Suadi" = c(1,1,1,1,1,1,1,1,0,1,1,1,0,0,0,1,1,1,1)
Krebs %v% "Pilot" = c(0,1,0,0,0,0,0,0,1,0,0,0,0,1,1,0,0,0,0)
Krebs %v% "alldeg" = degree(Krebs)

# set up
library(igraph)
library(intergraph)

# converts into igraph object
krebs_graph<-asIgraph(Krebs)
# extract data frames with node and edge info
krebs.v.df = as_data_frame(krebs_graph,what="vertices")
krebs.e.df = as_data_frame(krebs_graph,what="edges")
# write CSV files 
write.csv(krebs.v.df,file="krebs_v.csv") #Vertices 
write.csv(krebs.e.df,file="krebs_e.csv") #Edges

```


## Visualization and Network Summary:
Review some characteristics of the network
```{r, message=FALSE}

# switch libraries
detach(package:igraph)
library(intergraph)
library(statnet)

# get network size
network.size(Krebs)

# gey density
gden(Krebs)

```

Density is the proportion of observed ties (also called edges, arcs, or relations) in a network to the maximum number of possible ties. Thus, density is a ratio that can range from 0 to 1. The closer to 1 the density is, the more interconnected is the network. We find here a density of 0.15789. This shows a lower proportion of observed ties. We find a network size of 19. 


```{r}
# get diameter
lgc <- component.largest(Krebs,result="graph") 
gd <- geodist(lgc) 
max(gd$gdist)

```

The diameter of a network is a useful measure of this compactness. A path is the series of steps required to go from node A to node B in a network. The shortest path is the shortest number of steps required. The diameter then for an entire network is the longest of the shortest paths across all pairs of nodes. This is a measure of compactness or network efficiency in that the diameter reflects the ‘worst case scenario’ for sending information (or any other resource) across a network.

```{r, warning=FALSE}
# get clustering/transitivity 
gtrans(Krebs,mode="graph")

```

We find clustering or transitivity to be the proportion of closed triangles (triads where all three ties are observed) to the total number of open and closed triangles (triads where either two or all three ties are observed). Thus, like density, transitivity is a ratio that can range from 0 to 1.

Lets start to visualize the network and see if we are able to optimize that visualization.

```{r}
# plot the network using defaults
gplot(Krebs, vertex.col = 2, displaylabels = TRUE)

```

This plot provides the defualt visualization from the networks package. We can examine some ways to improve this visualization

```{r}
# take a look at the attributes in the network object
list.vertex.attributes(Krebs)

# review what those attributes contain
get.vertex.attribute(Krebs, "flight")
get.vertex.attribute(Krebs, "name")

```

Examine different methods to plot the network 

```{r}

op <- par(mar = rep(0, 4),mfrow=c(1,2)) 
# produce the plot using the circle method
plot(Krebs,mode="circle",vertex.cex=1.5, main="Circle Method")
# produce the plot using the Fruchterman-Reingold algorithm
plot(Krebs,mode="fruchtermanreingold",vertex.cex=1.5, main="Fruchterman-Reingold") 
par(op)


```

We find here that utilizing the Fruchterman-Reingold algorithm to plot the network illustrates its entirety well. Let's now compare another plotting method against the more appropriate Fruchterman-Reingold algorithm

```{r}

op <- par(mar = c(0,0,4,0),mfrow=c(1,2)) 
gplot(Krebs,gmode="graph",mode="random", vertex.cex=1.5,main="Random layout")
gplot(Krebs,gmode="graph",mode="fruchtermanreingold", vertex.cex=1.5,main="Fruchterman-Reingold")
par(op)

```

We see from the examinations above that applying either attribute flight or name can improve the visualization of the network. Additionally, utilizing the Fruchterman-Reingold algorithm will assist in the understanding of the visualization.

A more important use of color is to communicate some characteristic of the node or network by having different nodes have different colors. Specifically, information stored in a categorical node attribute can often be communicated through judicious node color choices.

```{r}
# get list of attribute "flight" for plotting
flight_lab <- get.vertex.attribute(Krebs,"flight") 
# get list of names 
names_lab <- get.vertex.attribute(Krebs,"name") 
# set up output parameters
op <- par(mar=c(0,0,0,0)) 
# plot the krebs network with flights labeled and colored
plot(Krebs,usearrows=FALSE,vertex.cex=1.5,label=names_lab, displaylabels=T,vertex.col="flight")
par(op)

```

We see further improvement in the visualization. One final improvement can be in the color pallette used. We can use a custom pallete with the RColorBrewer library.

```{r}
library(RColorBrewer) 
# take a look at some avaialbe palletes. This will select 4 colors from the larger Set3 pallette
display.brewer.pal(4, "Set3")

```

Add the newly selected color pallette to the plot. To do so we must ensure the vertex is formatted as an factor. The indexing works with factors or numeric vectors, but not character vectors

```{r}
# create objet with chosen pallette
my_pal <- brewer.pal(4,"Set3") 
# set flight attribute as factor
flight_cat <- as.factor(get.vertex.attribute(Krebs,"flight"))
# plot network to review
plot(Krebs,vertex.cex=1.5,label=names_lab, displaylabels=T,vertex.col=my_pal[flight_cat])

```

Node size can also be used to improve the network visualization. We will calculate three different measures of node centrality to apply to the plot.

```{r}
# get node-level vector of quantitative information
# assign degree to an object
deg <- degree(Krebs,gmode="graph")
# assign closeness to an object
cls <- closeness(Krebs,gmode="graph") 
# assign betweenness to an object
bet <- betweenness(Krebs,gmode="graph") 
# compare plots for degree
op <- par(mar = c(0,0,2,1),mfrow=c(1,2)) 
# plot the "raw" values of deg
plot(Krebs,usearrows=T,vertex.cex=deg,main="Raw")
# plot the log of deg values
plot(Krebs,usearrows=FALSE,vertex.cex=log(deg), main="Adjusted")
par(op)

```

Now that we have determined the parameters we can apply to optimize the plot, the follwoing will produce that optimized visualization.

```{r}

# plot network with the color pallette, centrality information for node size, and a legend
plot(Krebs,vertex.cex=deg,label=names_lab, displaylabels=T, label.cex=0.5, pad=0.4,label.col="darkblue", vertex.col=my_pal[flight_cat])
legend("bottomleft",legend=c("AA-77","UA-93","UA-175","AA-11"),
col=my_pal,pch=19,pt.cex=1.0,bty="n", title="Flight Number", cex = 0.6)

```

We can see there are a few actors in the network that are slighty more powerful given there centrality information. We can see distinctive groups from each color indicating which flight and what actors interacted between those flights. Next, lets take a more advaced look at possible sub groups and communities.

```{r}
# determine which actor has the highest degree
# list names as reminder
Krebs %v% 'vertex.names'
# calcualte degree for each
degree(Krebs, gmode="graph")

```

We can see that actors Nawaf Alhazmi and Hamza Alghamdi have the highest degrees. This too is illustrated in the plot. Next we will take a look at closeness and betweenness.

```{r}
# calculate closeness for each 
closeness(Krebs, gmode="graph")
# calculate betweeness for each
betweenness(Krebs, gmode="graph")

```

We find the actor with the highest closeness to be Hamza Alghamdi. The actor with the highest betweeness is also Hamza Alghamdi followed closely behind by Nawaf Alhazmi. 

## Community Detection 

### Sub groups
Next, we will take a look at some of the simplest types of cohesive subgroups. Lets examine the network first for cliques.

```{r, message=FALSE}
# to do so we will need to switch over to the igraph package using the object we created previously
# switch libraries to avoid dependency conflicts
detach(package:statnet)
library(igraph)
library(intergraph)

# calcualte all cliques with a minimum of four
cliques(krebs_graph, min=4)

# calcualte  the largest clique
largest.cliques(krebs_graph)

```

We find the largest clique within the network to be Nawaf Alhazmi, Ahmed Alnami, Saeed Alghamdi, and Hamza Alghamdi. This subgroup contains the induviduals indentified earlier whilst examining betweeness and closeness. As cliques are useful here, they simply are not very common in larger social networks. As such, we will explore others methods to determine communities.

### Communities

```{r}
# assign an object representing clusters from the walk trap algorithm
cw <- cluster_walktrap(krebs_graph) 
# get membership
membership(cw)
# get modularity
modularity(cw)

```

We will consider the modularity high and we can see from the membership calculation that theres was found to be four different sub groups. A plot of this information will help illustrate.

```{r}
# plot communities
 plot(cw, krebs_graph)

```

We see here the distinct subgroups found by the cluster walk trap algorithm. Lets examine some additional methods.

```{r}
#Springlass
cs <- cluster_spinglass(krebs_graph) 
modularity(cs)
membership(cs)
#Fast-greedy
cfg <- cluster_fast_greedy(krebs_graph)
modularity(cfg) 
membership(cfg)
#Leading eigenvector
cle <- cluster_leading_eigen(krebs_graph)
modularity(cle)
membership(cle)

```

These results show that all the detection algorithms identify either two or five subgroups. Modularity ranges from about 0.46 to 0.50. Lets take a look at the plots from each of these algorithms.

```{r, fig.width=5, fig.height=5}
op <- par(mfrow=c(2,2),mar=c(2,0,2,0)) 
plot(cw, krebs_graph,vertex.label=V(krebs_graph)$flight,
main="Walktrap")
plot(cs, krebs_graph,vertex.label=V(krebs_graph)$flight,
main="Springlass")
plot(cfg, krebs_graph,vertex.label=V(krebs_graph)$flight,
main="Fast-greedy")
plot(cle, krebs_graph,vertex.label=V(krebs_graph)$flight,
main="Leading Eigenvector")

```

The plot above illustrates well the difference in the groups found by the algorithm. The nodes were labeled by flight as that would be assumed to be one possible manner of forming groups. We can see that not all groups are formed by flights though. We do see the groups of four and five formed by each. 

## Network Modeling

Exponential-family random graph models will be used to further analyze the network.
```{r, message=FALSE}
# to do so we will need to switch over to the statnet package using the object we created previously
# switch libraries to avoid dependency conflicts
detach(package:igraph)
library(statnet)
library(intergraph)

# fit ergm model to see what impact the flight attribute has on connections in the network
library(ergmharris)
null_model <-ergm(Krebs ~ edges)
summary(null_model)

```

Lets see how the model with only edges performed. The following will simulate the network from the model.

```{r}
# set up simulation
simtrinull<-simulate( null_model, nsim = 100, monitor=~triangles, statsonly=TRUE, control=control.simulate.ergm(MCMC.burnin=1000, MCMC.interval=1000), seed=567)
krebs.tri <- summary( Krebs~triangle ) 
#dev.off()
# plot results
par( mar = c( 4,4,1,1 ), cex.main = .9, cex.lab = .9,cex.axis = .75 )
hist(simtrinull[,2], xlim=c(0,9), col='coral', main="", xlab="Number of triangles", ylab="Number of simulations") 
points(krebs.tri,3, pch="X", cex=2)

```

After 100 simulations of the network based on the null model, we find that the null model did not quite capture the amount of triangles in the network, which was nine. Lets take a look at some more complex models to see if we can find a better fit.

Lets compare some models to see if we can find significant variables that assist in predicting the connection between two terrorist “nodes”. 

It is the hypothesis herein that age and whether or not one is a pilot are significant variables in predicting the connection between two terrorist “nodes”. 

```{r, warning=FALSE, message=FALSE}
# fit model and summarize
age_effects <-ergm(Krebs ~edges +nodecov('Age'))
summary(age_effects)

```


```{r, message=FALSE}
# fit model and summarize
age_origin_effects <-ergm(Krebs ~edges +nodecov('Age') +nodecov('Suadi'))
summary(age_origin_effects)

```


```{r, message=FALSE}
# fit model and summarize
age_origin_pilot_effects <-ergm(Krebs ~edges +nodecov('Age') +nodecov('Suadi') +nodecov('Pilot'))
summary(age_origin_pilot_effects)

```

```{r, message=FALSE}
# fit model and summarize
pilot_Age_effects <-ergm(Krebs ~edges +nodecov('Pilot') +nodecov('Age'))
summary(pilot_Age_effects)

```

We do not find performance enhancements thru any iteration of the model as exhibited by the increasing AIC in each successive model. Lets look further by fitting new models and running simulations.

```{r, message=FALSE}
# fit model and summarize
null_model <-ergm(Krebs ~ edges) #Null Model
nullsim <- simulate(null_model, verbose = TRUE,seed = 5) #Simulations based on Null model
mainsim <- simulate(pilot_Age_effects, verbose = TRUE, seed = 5) #Simulations based on pilot and age effect model
rowgof <- rbind(summary(Krebs ~ edges + degree(0:5) + triangle),
summary(nullsim ~ edges + degree(0:5) + triangle), summary(mainsim ~ edges + degree(0:5) + triangle))
rownames(rowgof) <- c("Krebs", "Null", "Pilot Age effects")
rowgof

```

We do find the pilot age effect model capturing more edges than the null model. Similarly, we fnd that it captures more triangles as well. Next lets take a look more at how the model did with goodness of fit.

```{r, warning=FALSE, message=FALSE}

diff2_gof <- gof( pilot_Age_effects, GOF = ~degree + espartners + dspartners, verbose = T, burnin = 10000, interval = 10000 )
diff2_gof

```


```{r, fig.height=8, fig.width=8}
# plot results
par( mfrow = c( 3,2 ) )
plot( diff2_gof, cex.lab = 1.5, cex.axis = 1.5 ) 
plot(diff2_gof, cex.lab = 1.5, cex.axis = 1.5, plotlogodds = T )


```
Illustrated here is model performance, which can see is acceptable for degree. The model captured well degree and dyad wise shared partners. Where we can see it has had some issues is with capturing edge wise shared partners. 



























